Eco-evolutionary dynamics of finite populations from first principles

### 2023-02-06 ind_files/03_BD_processes.tex
Make notation consistent across multidimensional and field eqns

b^{(K)}_i(\mathbf{x}) &= \mu Q_i(\mathbf{x}) + b^{(\textrm{ind})}_{i}(\mathbf{x})x_i\\
d^{(K)}_i(\mathbf{x}) &= d^{(\textrm{ind})}_i(\mathbf{x})x_i

@@

where $\mu \geq 0$ is a constant describing the mutation rate, $Q_i(\mathbf{x})$ is a real-valued function that describes the birth rate of type $i$ individuals due to mutations in the population $\mathbf{x}$, and $b^{(\textrm{ind})}_{i}(\mathbf{x})$ and $d^{(\textrm{ind})}_{i}(\mathbf{x})$ are real-valued functions that respectively describe the per-capita birth and death rate of type $i$ individuals. Our assumptions of the functional forms \eqref{nD_functional_forms_for_replicator} thus amount to saying that birth and death rates can be separated into mutational and non-mutational components, and furthermore that the density dependence of the birth and death rates due to non-mutational effects is in a form that allows us to write down per-capita birth and death rates. Plugging these definitions into \eqref{nD_det_limit} and writing it down component-wise, we obtain the equation

@@

Where for each $i$, we have defined the real-valued function $w_i(\mathbf{x}) \coloneqq b^{(\textrm{int})}_{i}(\mathbf{x}) - d^{(\textrm{int})}_i(\mathbf{x})$. Setting the mutation rate $\mu \to 0$ in equation \eqref{nD_det_limit_fitess_defn} makes it clear that the quantity $w_i(\mathbf{x})x_i$ describes the growth rate of type $i$ individuals in the population due to all processes other than mutation. The quantity $w_i(\mathbf{x})$ thus describes the per-capita growth rate of type $i$ individuals in a population $\mathbf{x}$, and is sometimes called the `Malthusian fitness' of type $i$. Ecologists often denote this quantity by the symbol $r_i$ and simply call it the (exponential) growth rate of type $i$, but we will stick to $w_i$ and `fitness' here. It is notable that the fitness of a type depends on the state of the population as a whole (\textit{i.e.} $\mathbf{x}$) and is thus frequency-dependent.\\

@@

Following ideas similar in spirit to \citep{mcleod_social_2019}, we can also use \eqref{nD_Ito_SDE} to write down stochastic equations for $p_i(t)$, the frequency of type $i$ individuals in the population to contrast with the replicator-mutator equation \eqref{nD_replicator_mutator} derived above. In appendix \ref{App_density_to_freq}, we show that we can use It\^{o}'s formula to write  down a general stochastic equation for the frequencies of each type in the population. Unlike \citep{mcleod_social_2019}, we make no assumptions about the separation of ecological and evolutionary time scales or the strength of selection and are able to present an entirely general calculation. Letting $\tau_i(\mathbf{x}) = b^{(\textrm{ind})}_{i}(\mathbf{x}) + d^{(\textrm{ind})}_i(\mathbf{x})$ be the \emph{per-capita turnover rate} of the $i\textsuperscript{th}$ type and denoting $\overline{\tau} = \sum_i p_i\tau_i$, we show in appendix \ref{App_density_to_freq} that when the birth and death rates are defined by \eqref{nD_functional_forms_for_replicator}, then the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{X}_t$, changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\left[\left(A^{+}_{i}\right)^{1/2}dB^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dB^{(j)}_t\right]
\end{aligned}
\end{equation}
where $B^{(1)}_t,B^{(2)}_t, \ldots, B^{(m)}_t$ are $m$ independent one-dimensional standard Brownian motion processes. The first term in this expression is the same as equation \eqref{nD_replicator_mutator} and describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation. We saw in the previous section that the replicator-mutator equation is generically satisfied in the infinite population limit for all birth-death processes that are in the (very general) functional forms \eqref{nD_functional_forms_for_replicator}, a result that is made explicit in \eqref{nD_eqn_for_frequencies} (just take $K \to \infty$ to convince yourself). The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, along with the connection to previous studies and standard infinite population models, are discussed in chapter \ref{chap_unification}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect'. To the highest order in $\sqrt{K}$, the effect of these fluctuations can be described through the weak noise approximation, as we explain below. However, the $1/K$ factor in the second term of \eqref{nD_eqn_for_frequencies} means that the effect of differential turnover rates disappears in this weak noise regime  (and also in other such approximate treatments, such as the Chemical Langevin equation).

>>>>

b^{(K)}_i(\mathbf{x}) &= \mu Q_i(\mathbf{x}) + x_ib^{\textrm{(ind)}}_{i}(\mathbf{x})\\
d^{(K)}_i(\mathbf{x}) &= x_id^{\textrm{(ind)}}_i(\mathbf{x})

@@

where $\mu \geq 0$ is a constant describing the mutation rate, $Q_i(\mathbf{x})$ is a real-valued function that describes the birth rate of type $i$ individuals due to mutations in the population $\mathbf{x}$, and $b^{\textrm{(ind)}}_{i}(\mathbf{x})$ and $d^{\textrm{(ind)}}_{i}(\mathbf{x})$ are real-valued functions that respectively describe the per-capita birth and death rate of type $i$ individuals. Our assumptions of the functional forms \eqref{nD_functional_forms_for_replicator} thus amount to saying that birth and death rates can be separated into mutational and non-mutational components, and furthermore that the density dependence of the birth and death rates due to non-mutational effects is in a form that allows us to write down per-capita birth and death rates. Plugging these definitions into \eqref{nD_det_limit} and writing it down component-wise, we obtain the equation

@@

Where for each $i$, we have defined the real-valued function $w_i(\mathbf{x}) \coloneqq b^{\textrm{(ind)}}_{i}(\mathbf{x}) - d^{\textrm{(ind)}}_{i}(\mathbf{x})$. Setting the mutation rate $\mu \to 0$ in equation \eqref{nD_det_limit_fitess_defn} makes it clear that the quantity $w_i(\mathbf{x})x_i$ describes the growth rate of type $i$ individuals in the population due to all processes other than mutation. The quantity $w_i(\mathbf{x})$ thus describes the per-capita growth rate of type $i$ individuals in a population $\mathbf{x}$, and is sometimes called the `Malthusian fitness' of type $i$. Ecologists often denote this quantity by the symbol $r_i$ and simply call it the (exponential) growth rate of type $i$, but we will stick to $w_i$ and `fitness' here. It is notable that the fitness of a type depends on the state of the population as a whole (\textit{i.e.} $\mathbf{x}$) and is thus frequency-dependent.\\

@@

Following ideas similar in spirit to \citep{mcleod_social_2019}, we can also use \eqref{nD_Ito_SDE} to write down stochastic equations for $p_i(t)$, the frequency of type $i$ individuals in the population to contrast with the replicator-mutator equation \eqref{nD_replicator_mutator} derived above. In appendix \ref{App_density_to_freq}, we show that we can use It\^{o}'s formula to write  down a general stochastic equation for the frequencies of each type in the population. Unlike \citep{mcleod_social_2019}, we make no assumptions about the separation of ecological and evolutionary time scales or the strength of selection and are able to present an entirely general calculation. Letting $\tau_i(\mathbf{x}) = b^{\textrm{(ind)}}_{i}(\mathbf{x}) + b^{\textrm{(ind)}}_{i}(\mathbf{x})$ be the \emph{per-capita turnover rate} of the $i\textsuperscript{th}$ type and denoting $\overline{\tau} = \sum_i p_i\tau_i$, we show in appendix \ref{App_density_to_freq} that when the birth and death rates are defined by \eqref{nD_functional_forms_for_replicator}, then the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{X}_t$, changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\left[\left(A^{+}_{i}\right)^{1/2}dB^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dB^{(j)}_t\right]
\end{aligned}
\end{equation}
where $B^{(1)}_t,B^{(2)}_t, \ldots, B^{(m)}_t$ are $m$ independent one-dimensional standard Brownian motion processes. The first term in this expression is the same as equation \eqref{nD_replicator_mutator} and describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation. We saw in the previous section that the replicator-mutator equation is generically satisfied in the infinite population limit for all birth-death processes that are in the (very general) functional forms \eqref{nD_functional_forms_for_replicator}, a result that is made explicit in \eqref{nD_eqn_for_frequencies} (just take $K \to \infty$ to convince yourself). The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, along with the connection to previous studies and standard infinite population models, are discussed in chapter \ref{chap_unification}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect'. To the highest order in $\sqrt{K}$, the effect of these fluctuations can be described through the weak noise approximation, as we explain below. However, the $1/K$ factor in the second term of \eqref{nD_eqn_for_frequencies} means that the effect of differential turnover rates disappears in this weak noise regime  (and also in other such approximate treatments, such as the Chemical Langevin equation).

### 2023-02-06 ind_files/04_field_eqns.tex
Make notation consistent across multidimensional and field eqns

If you know and care about enough mathematics for this to really bother you, see \citep{champagnat_unifying_2006} for a much more rigorous treatment that avoids using informal tools such as functional derivatives and functional equivalents of Fokker-Planck equations in favor of a probabilistic approach grounded in (measure-theoretic) Markov and martingale theory.}. The state space of the stochastic process thus becomes infinite-dimensional, which complicates matters slightly. The principal objects of interest here are \emph{functionals} $F[x, \phi^{(t)}]$ which take in a scalar $x$ representing the trait value of interest, and a function $\phi^{(t)}$ representing the population at time $t$. Thus, whereas in the previous section we were interested in how a function $f(x(t))$ changes based on the change in an input variable $x(t)$ (the population), we are now interested in how a functional $F[\phi^{(t)}]$ changes with the change in an input function $\phi^{(t)}$. The mathematics for these sorts of processes is an active area of research and is comparatively far from well developed. The mathematically rigorous formulation of the kinds of processes we study here falls in the realm of measure-valued branching processes, and is highly technical and rather inaccessible unless one is already comfortable with advanced measure-theoretic notions \citep{champagnat_unifying_2006,champagnat_individual_2008}. This means that the existing formalism, while admirable in its generality and mathematical rigor, is rather unusable for most biologists, who do not have formal training in analysis. One can, however, make progress if they are willing to take some mathematical leaps of faith and sacrifice rigor for the sake of accessibility and heuristic understanding. I will shamelessly adopt this attitude below and hope that any questions of rigor, well-posedness, existence, etc. will be sorted out by some clever mathematicians in the future. Physicists use the term `field' for functions of the form $f(x,t):\mathbb{R}^{n} \times [0,\infty) \to \mathbb{R}^{m}$, where $\mathbb{R}^{n}$ represents space and $[0,\infty)$ represents time. They then call models which describe such functions `field theories'. In physics jargon, the stochastic process we will formulate $\{\phi^{(t)}\}_{t \geq 0}$ when viewed as a sequence of functions $\{\phi^{(t)}(y)\}_{t \geq 0}$ thus describes a (scalar) `stochastic field', and the formalism we will develop below is a `stochastic field theory' of evolution, where physical space has been replaced by an abstract trait space. This is closely related to the area of physics called `statistical field theory', the analog of quantum field theory for systems with a large number of classical particles. Stochastic field theories over physical space have recently been used in biology to model brain function \citep{bressloff_stochastic_2010} and collective motion \citep{o_laighleis_minimal_2018}. In the following sections, I will rely heavily on a heuristic object called the functional derivative $\delta F/\delta \phi$. The functional derivative is an \emph{ad hoc}, somewhat informal notion, defined indirectly as the unique object that obeys, for any `nice' function $\rho$

@@

b_K(x|\psi) &= \mu b_{\textrm{mut}}(x|\psi) + b_{\textrm{ind}}(x|\psi)\psi(x,t)\\
d_K(x|\psi) &= d_{\textrm{ind}}(x|\psi)\psi(x,t)

@@

Here, $ b_{\textrm{mut}}(x|\psi)$ describes birth due to mutations, and $\mu \geq 0$ is a mutation rate. The functions $b_{\textrm{ind}}(x|\psi)$ and $d_{\textrm{ind}}(x|\psi)$ describe the per-capita birth rate and death rate of type $x$ in a population $\psi$. For example,  $b_{\textrm{ind}}(x|\psi)$ may incorporate the effects of mate choice in the sexual case or intrinsic duplication rate in the asexual case, and $d_{\textrm{int}}(x|\psi)$ may model death due to intraspecific competition for resources. Substituting equation \eqref{BD_for_kimura} into \eqref{deterministic_traj}, we obtain

@@

\frac{\partial \psi}{\partial t}(x,t) = w(x|\psi)\psi(x,t) + \mu b_{\textrm{mut}}(x|\psi)

@@

where we have defined $w(x|\psi) \coloneqq b_{\textrm{int}}(x|\psi) - d_{\textrm{int}}(x|\psi)$, which can be thought of as the (Malthusian) `fitness' of the phenotype $x$. To track population numbers and trait frequencies, we follow the approach of \citep{week_white_2021} and define

@@

\frac{\partial p}{\partial t} &= \frac{1}{N_K(t)}\left[w(x|\psi)\psi(x,t) + \mu b_{\textrm{mut}}(x|\psi)\right] - \frac{\psi(x,t)}{N^2_K(t)}\int\limits_{\mathcal{T}}w(y|\psi)\psi(y,t) + \mu b_{\textrm{mut}}(y|\psi)dy\\
&= w(x|\psi)p(x,t) + \frac{\mu}{N_{K}(t)} b_{\textrm{mut}}(x|\psi) - p(x,t)\left(\int\limits_{\mathcal{T}}w(y|\psi)p(y,t)dy+\frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}} b_{\textrm{mut}}(y|\psi)dy\right)

@@

\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial p}{\partial t}(x,t) = \left[w(x|\psi) - \overline{w}(t)\right]p(x,t)+\frac{\mu}{N_K(t)}\left[b_{\textrm{mut}}(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} b_{\textrm{mut}}(y|\psi)dy\right]}

@@

This is a continuous version of the replicator-mutator equation when each $x$ is viewed as a strategy. It also yields Kimura's continuum-of-alleles model when each $x$ is viewed as an allele, $b_{\textrm{mut}}(x|\psi)$ takes the form of a convolution of $\psi(x,t)$ with a mutation kernel, and the trait space is the entire real line, \emph{i.e.} $\mathcal{T} = \mathbb{R}$. To see this, let $b_{\textrm{mut}}(y|\psi) = \int\limits_{\mathbb{R}}m(y-z)\psi(z,t)dz$, where $m(x)$ is a mutation kernel, which by definition is normalized such that $\int\limits_{\mathbb{R}}m(x)dx = 1$. Let us further note that we have implicitly been assuming that the total number of individuals (scaled by $K$) remains finite at all times, \emph{i.e.} $N_K(t) = \int\limits_{\mathbb

@@

\int\limits_{\mathbb{R}} b_{\textrm{mut}}(y|\psi)dy &= \int\limits_{\mathbb{R}}\int\limits_{\mathbb{R}} m(y-z)\psi(z,t)dzdy\nonumber\\

@@

where we have used the Fubini-Tonnelli theorem to go from the first step to the second, and have made the substitution $u = y-z$ to go from the third to the fourth step. We then note that since $m$ is a kernel, it satisfies $\int\limits_{\mathbb{R}} m(u)du = 1$, and \eqref{convolution_integral_intermediate} therefore becomes $\int\limits_{\mathbb{R}} b_{\textrm{mut}}(y|\psi)dy = N_K(t)$. Substituting this in \eqref{cts_replicator_mutator}, we have

@@

\frac{d \overline{x}}{dt} &= \int\limits_{\mathcal{T}}xw(x|\psi)p(x,t)dx - \overline{w}(t)\int\limits_{\mathcal{T}}xp(x,t)dx+\frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}}x\left[b_{\textrm{mut}}(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} b_{\textrm{mut}}(y|\psi)dy\right]dx\nonumber\\
&= \overline{xw} - \overline{w}\cdot\overline{x} + \frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}}x\left[b_{\textrm{mut}}(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} b_{\textrm{mut}}(y|\psi)dy\right]dx\label{intermediate_for_cts_price}

@@

M(x|\psi) \coloneqq \frac{\mu}{N_K(t)}\left[\int\limits_{\mathcal{T}}xb_{\textrm{mut}}(x|\psi)dx - \left(\overline{x}\int\limits_{\mathcal{T}}b_{\textrm{mut}}(x|\psi)dx\right)\right] 

@@

    \item Small mutational effects with `almost faithful' reproduction, meaning $b_{\textrm{mut}}(x|\psi) \to 0$, and the distribution $\psi(x,t)$ tends to stay very `sharp' (i.e strongly peaked about its mean value). 

>>>>

If you know and care about enough mathematics for this to really bother you, see \citep{champagnat_unifying_2006} for a much more rigorous treatment that avoids using informal tools such as functional derivatives and functional equivalents of Fokker-Planck equations in favor of a probabilistic approach grounded in (measure-theoretic) Markov and martingale theory.}. The state space of the stochastic process thus becomes infinite-dimensional, which complicates matters slightly. The principal objects of interest here are \emph{functionals} $F[x, \phi^{(t)}]$ which take in a scalar $x$ representing the trait value of interest, and a function $\phi^{(t)}$ representing the population at time $t$. Thus, whereas in the previous section we were interested in how a function $f(x(t))$ changes based on the change in an input variable $x(t)$ (the population), we are now interested in how a functional $F[\phi^{(t)}]$ changes with the change in an input function $\phi^{(t)}$. The mathematics for these sorts of processes is an active area of research and is comparatively far from well developed. The mathematically rigorous formulation of the kinds of processes we study here falls in the realm of measure-valued branching processes, and is highly technical and rather inaccessible unless one is already comfortable with advanced measure-theoretic notions \citep{champagnat_unifying_2006,champagnat_individual_2008}. This means that the existing formalism, while admirable in its generality and mathematical rigor, is rather unusable for most biologists, who do not have formal training in analysis (but see \cite{week_white_2021} for a very friendly introduction to the major ideas through heuristics). One can, however, make progress if they are willing to take some mathematical leaps of faith and sacrifice rigor for the sake of accessibility and heuristic understanding. I will shamelessly adopt this attitude below and hope that any questions of rigor, well-posedness, existence, etc. will be sorted out by some clever mathematicians in the future. Physicists use the term `field' for functions of the form $f(x,t):\mathbb{R}^{n} \times [0,\infty) \to \mathbb{R}^{m}$, where $\mathbb{R}^{n}$ represents space and $[0,\infty)$ represents time. They then call models which describe such functions `field theories'. In physics jargon, the stochastic process we will formulate $\{\phi^{(t)}\}_{t \geq 0}$ when viewed as a sequence of functions $\{\phi^{(t)}(y)\}_{t \geq 0}$ thus describes a (scalar) `stochastic field', and the formalism we will develop below is a `stochastic field theory' of evolution, where physical space has been replaced by an abstract trait space. This is closely related to the area of physics called `statistical field theory', the analog of quantum field theory for systems with a large number of classical particles. Stochastic field theories over physical space have recently been used in biology to model brain function \citep{bressloff_stochastic_2010} and collective motion \citep{o_laighleis_minimal_2018}. In the following sections, I will rely heavily on a heuristic object called the functional derivative $\delta F/\delta \phi$. The functional derivative is an \emph{ad hoc}, somewhat informal notion, defined indirectly as the unique object that obeys, for any `nice' function $\rho$

@@

b_K(x|\psi) &= \mu b^{\textrm{(mut)}}(x|\psi) + \psi(x,t)b^{\textrm{(ind)}}(x|\psi)\\
d_K(x|\psi) &= \psi(x,t)d^{\textrm{(ind)}}(x|\psi)

@@

Here, $ b^{\textrm{(mut)}}(x|\psi)$ describes birth due to mutations, and $\mu \geq 0$ is a mutation rate. The functions $b^{\textrm{(ind)}}(x|\psi)$ and $d^{\textrm{(ind)}}(x|\psi)$ describe the per-capita birth rate and death rate of type $x$ in a population $\psi$. These functions could in principle model several ecological factors. For example,  $b^{\textrm{(ind)}}(x|\psi)$ may incorporate the effects of mate choice in the sexual case or intrinsic duplication rates in the asexual case, and $d^{\textrm{(ind)}}(x|\psi)$ may model death due to intraspecific competition for resources. Substituting equation \eqref{BD_for_kimura} into \eqref{deterministic_traj}, we obtain

@@

\frac{\partial \psi}{\partial t}(x,t) = w(x|\psi)\psi(x,t) + \mu b^{\textrm{(mut)}}(x|\psi)

@@

where we have defined $w(x|\psi) \coloneqq b^{\textrm{(ind)}}(x|\psi) - d^{\textrm{(ind)}}(x|\psi)$, which can be thought of as the (Malthusian) `fitness' of the phenotype $x$. To track population numbers and trait frequencies, we follow the approach of \citep{week_white_2021} and define

@@

\frac{\partial p}{\partial t} &= \frac{1}{N_K(t)}\left[w(x|\psi)\psi(x,t) + \mu b^{\textrm{(mut)}}(x|\psi)\right] - \frac{\psi(x,t)}{N^2_K(t)}\int\limits_{\mathcal{T}}w(y|\psi)\psi(y,t) + \mu b^{\textrm{(mut)}}(y|\psi)dy\\
&= w(x|\psi)p(x,t) + \frac{\mu}{N_{K}(t)} b^{\textrm{(mut)}}(x|\psi) - p(x,t)\left(\int\limits_{\mathcal{T}}w(y|\psi)p(y,t)dy+\frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}} b^{\textrm{(mut)}}(y|\psi)dy\right)

@@

\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial p}{\partial t}(x,t) = \left[w(x|\psi) - \overline{w}(t)\right]p(x,t)+\frac{\mu}{N_K(t)}\left[b^{\textrm{(mut)}}(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} b^{\textrm{(mut)}}(y|\psi)dy\right]}

@@

This is a continuous version of the replicator-mutator equation when each $x$ is viewed as a strategy. It also yields Kimura's continuum-of-alleles model when each $x$ is viewed as an allele, $b^{\textrm{(mut)}}(x|\psi)$ takes the form of a convolution of $\psi(x,t)$ with a mutation kernel, and the trait space is the entire real line, \emph{i.e.} $\mathcal{T} = \mathbb{R}$. To see this, let $b^{\textrm{(mut)}}(y|\psi) = \int\limits_{\mathbb{R}}m(y-z)\psi(z,t)dz$, where $m(x)$ is a mutation kernel, which by definition is normalized such that $\int\limits_{\mathbb{R}}m(x)dx = 1$. Let us further note that we have implicitly been assuming that the total number of individuals (scaled by $K$) remains finite at all times, \emph{i.e.} $N_K(t) = \int\limits_{\mathbb

@@

\int\limits_{\mathbb{R}} b^{\textrm{(mut)}}(y|\psi)dy &= \int\limits_{\mathbb{R}}\int\limits_{\mathbb{R}} m(y-z)\psi(z,t)dzdy\nonumber\\

@@

where we have used the Fubini-Tonnelli theorem to go from the first step to the second, and have made the substitution $u = y-z$ to go from the third to the fourth step. We then note that since $m$ is a kernel, it satisfies $\int\limits_{\mathbb{R}} m(u)du = 1$, and \eqref{convolution_integral_intermediate} therefore becomes $\int\limits_{\mathbb{R}} b^{\textrm{(mut)}}(y|\psi)dy = N_K(t)$. Substituting this in \eqref{cts_replicator_mutator}, we have

@@

\frac{d \overline{x}}{dt} &= \int\limits_{\mathcal{T}}xw(x|\psi)p(x,t)dx - \overline{w}(t)\int\limits_{\mathcal{T}}xp(x,t)dx+\frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}}x\left[b^{\textrm{(mut)}}(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} b^{\textrm{(mut)}}(y|\psi)dy\right]dx\nonumber\\
&= \overline{xw} - \overline{w}\cdot\overline{x} + \frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}}x\left[b^{\textrm{(mut)}}(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} b^{\textrm{(mut)}}(y|\psi)dy\right]dx\label{intermediate_for_cts_price}

@@

M(x|\psi) \coloneqq \frac{\mu}{N_K(t)}\left[\int\limits_{\mathcal{T}}xb^{\textrm{(mut)}}(x|\psi)dx - \left(\overline{x}\int\limits_{\mathcal{T}}b^{\textrm{(mut)}}(x|\psi)dx\right)\right] 

@@

    \item Small mutational effects with `almost faithful' reproduction, meaning $b^{\textrm{(mut)}}(x|\psi) \to 0$, and the distribution $\psi(x,t)$ tends to stay very `sharp' (i.e strongly peaked about its mean value). 

### 2023-02-06 ind_files/06_unification_discussion.tex
Make notation consistent across multidimensional and field eqns

where all covariances are understood in the statistical sense (Note that since each $w_i$, $\tau_i$, $\overline{w}$, and $\overline{\tau}$ are stochastic processes depending on $\mathbf{p}$, the terms $ \textrm{Cov}(w,f)$ and $\textrm{Cov}(\tau,f)$ are themselves stochastic processes). Taking $K \to \infty$ in equation \eqref{nD_stochastic_Price} recovers the standard Price equation as the infinite population limit. In some sense, \eqref{nD_stochastic_RM} and \eqref{nD_stochastic_Price} are therefore `fundamental' equations for the evolution of finite populations, with the replicator-mutator and Price equations as the infinite population limits respectively. In both equations, the first term represents the direct effects of forces captured in classic deterministic models - the selection and mutation are partitioned into two terms in \eqref{nD_stochastic_Price}, but since mutation acts symmetrically in both the first and second term of \eqref{nD_stochastic_RM}, this has no major bearing on our interpretations. As is evident in both \eqref{nD_stochastic_RM} and \eqref{nD_stochastic_Price}, finite populations experience a new directional force dependent on $\tau_i = b^{(ind)}_i + d^{(ind)}_i$, the per-capita turnover rate of type $i$, that cannot be captured in infinite population models. Remarkably, this term acts in a way that is mathematically exactly analogously to the classical evolutionary force of selection and mutation as captured by the first term in \eqref{nD_stochastic_RM}, but in the opposite direction - A higher relative $\tau_i$ leads to a decrease in frequency (Notice the minus sign before the second term in \eqref{nD_stochastic_RM}). Biologically, this term describes a biasing effect due to differential turnover rates and can intuitively be understood as being similar to the gambler's ruin through the following reasoning: If a type $i$ has a higher $\tau_i$, it experiences greater turnover due to a generally higher birth and death rate and thus experience more births and deaths in a given time interval than an otherwise equivalent species with a lower $\tau_i$. More events mean greater demographic stochasticity, and types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period. This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, which is reflected in the $1/N_K$ factor in this term. This stochastic analog of selection for reduced turnover rates, captured by the second term of equation \eqref{nD_stochastic_RM}, is the force responsible for the `reversal of the direction of deterministic selection' induced by demographic noise in previous studies \citep{houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015, constable_demographic_2016, mcleod_social_2019}. Note that types that tend to increase the \emph{total} population size $N_K(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will reduce the magnitude of this effect compared to types that do not facilitate such an increase, such as cheaters and highly competitive species, which could explain why this effect preferentially favors the former types in reversing the direction of deterministic selection in finite population models \citep{houchmandzadeh_fluctuation_2015, mcleod_social_2019}. The fact that total population size controls the strength of the second term of \eqref{nD_stochastic_RM} also explains why cooperation is favoured in the early transient period of population growth \citep{melbinger_evolutionary_2010} when simulations are initiated from a small population size - In the early transient period, $N_K(t)$ is small, and the biasing effect of differential turnover rates is stronger, thus favouring cooperation. More generally, selection for reduced turnover rate could explain why cooperation often persists for a long time in finite population IbMs (and the real world) despite infinite population models predicting their extinction. The fact that the entire term is multiplied by $(KN_K(t))^{-1}$ suggests that the effect of this force is weak for medium to large populations, which explains why the persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, rapid attraction to a slow manifold, or a weak selection + weak mutation limit. In all three of these cases, the first term on the RHS of \eqref{nD_eqn_for_frequencies} becomes identically 0. It therefore no longer contributes to the total dynamics and allows us to see the contributions of the second term.\\

>>>>

where all covariances are understood in the statistical sense (Note that since $w_i$, $\tau_i$, $\overline{w}$, and $\overline{\tau}$ are stochastic processes depending on $\mathbf{p}$, the terms $ \textrm{Cov}(w,f)$ and $\textrm{Cov}(\tau,f)$ are themselves stochastic processes). Taking $K \to \infty$ in equation \eqref{nD_stochastic_Price} recovers the standard Price equation as the infinite population limit. In some sense, \eqref{nD_stochastic_RM} and \eqref{nD_stochastic_Price} are therefore `fundamental' equations for the evolution of finite populations, with the replicator-mutator and Price equations as the infinite population limits respectively. In both equations, the first term represents the direct effects of forces captured in classic deterministic models - the selection and mutation are partitioned into two terms in \eqref{nD_stochastic_Price}, but since mutation acts symmetrically in both the first and second term of \eqref{nD_stochastic_RM}, this has no major bearing on our interpretations. As is evident in both \eqref{nD_stochastic_RM} and \eqref{nD_stochastic_Price}, finite populations experience a new directional force dependent on $\tau_i(\mathbf{x}) = b^{\textrm{(ind)}}_{i}(\mathbf{x}) + b^{\textrm{(ind)}}_{i}(\mathbf{x})$, the per-capita turnover rate of type $i$, that cannot be captured in infinite population models. Remarkably, this term acts in a way that is mathematically exactly analogously to the classical evolutionary force of selection and mutation as captured by the first term in \eqref{nD_stochastic_RM}, but in the opposite direction - A higher relative $\tau_i$ leads to a decrease in frequency (Notice the minus sign before the second term in \eqref{nD_stochastic_RM}). Biologically, this term describes a biasing effect due to differential turnover rates and can intuitively be understood as being similar to the gambler's ruin through the following reasoning: If a type $i$ has a higher $\tau_i$, it experiences greater turnover due to a generally higher birth and death rate and thus experience more births and deaths in a given time interval than an otherwise equivalent species with a lower $\tau_i$. More events mean greater demographic stochasticity, and types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period. This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, which is reflected in the $1/N_K$ factor in this term. This stochastic analog of selection for reduced turnover rates, captured by the second term of equation \eqref{nD_stochastic_RM}, is the force responsible for the `reversal of the direction of deterministic selection' induced by demographic noise in previous studies \citep{houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015, constable_demographic_2016, mcleod_social_2019}. Note that types that tend to increase the \emph{total} population size $N_K(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will reduce the magnitude of this effect compared to types that do not facilitate such an increase, such as cheaters and highly competitive species, which could explain why this effect preferentially favors the former types in reversing the direction of deterministic selection in finite population models \citep{houchmandzadeh_fluctuation_2015, mcleod_social_2019}. The fact that total population size controls the strength of the second term of \eqref{nD_stochastic_RM} also explains why cooperation is favoured in the early transient period of population growth \citep{melbinger_evolutionary_2010} when simulations are initiated from a small population size - In the early transient period, $N_K(t)$ is small, and the biasing effect of differential turnover rates is stronger, thus favouring cooperation. More generally, selection for reduced turnover rate could explain why cooperation often persists for a long time in finite population IbMs (and the real world) despite infinite population models predicting their extinction. The fact that the entire term is multiplied by $(KN_K(t))^{-1}$ suggests that the effect of this force is weak for medium to large populations, which explains why the persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, rapid attraction to a slow manifold, or a weak selection + weak mutation limit. In all three of these cases, the first term on the RHS of \eqref{nD_eqn_for_frequencies} becomes identically 0. It therefore no longer contributes to the total dynamics and allows us to see the contributions of the second term.\\
